\sysfull

### 2017-07-31 paper/background.tex
Corrected error in background.tex

\In this section, we provide background on energy harvesting systems and the
intermittent software execution model. We then discuss the key limitations of
existing task-based execution models for intermittent computing that \sys
addresses.

@@

Energy harvesting devices operate using energy extracted from their
environment, from sources such as radio waves (RF)~\cite{wisp} and solar
energy. Energy-harvesting devices need not use a tethered power supply or
battery and instead, typically collect energy into a (super-)capacitor, operate
when sufficient energy has accumulated, and upon depleting the stored energy,
turn off and wait to collect more energy. 
%
Batteryless operation has a number of important advantages, making intermittent
computing an important research domain. Supplying power to
billions~\cite{gartner_iot} of embedded computers using batteries is not
sustainable. The European Commission estimates that more than 160 kilotons of
consumer batteries enter the European Union annually~\cite{eu_batteries_2016}.
Batteries are an environmental risk, are fragile, are limited in their number
of charge/discharge cycles, and may require costly physical maintenance that is
difficult or impossible deployed (e.g. in space~\cite{kicksat}). By contrast,
super-capacitors are durable, promising millions of charge/discharge
cycles~\cite[Sec. I]{ongaro_pwre_2012}. The main limitation in moving to
capacitor-based energy storage is that capacitor energy density---and
consequently operating discharge time---is orders of magnitude less than a
battery.

%Given current technology development, battery-less systems are best suited for
%very long-term sensing and monitoring where access to recharge is either
%prohibitive or impossible. These include battery-less image capture and
%processing~\cite{naderiparizi_rfid_2015}, animal
%monitoring~\cite{thomas_jbcs_2012} or implantable~\cite{rodriguez_tbcs_2015}
%and digestible~\cite{nadeau_naturebio_2017} sensors.

@@

\subsection{Intermittent Computing}
\label{sec:background_consistency}
Energy-harvesting devices execute software according to the {\em intermittent
execution model}~\cite{dino,chain,alpaca,ratchet}.  Physically, a device
charges until a threshold, operates briefly until its energy is depleted, shuts
down recharges, and repeats the cycle.  Software running on the device executes
{\em intermittently} because energy is only available between when the
capacitor reaches its threshold and when it is depleted. An intermittent
execution is composed of operating periods interspersed with power failures.
The frequency of power failures in an intermittent execution depends on the
size of the device's energy storage buffer: a larger buffer allows longer
operating periods.  Energy-harvesters provide input power that is orders of
magnitude less than a device's operating power, making recharging negligible
during operation.

@@

Data may become inconsistent if an attempt to execute some computation {\em
writes} to a non-volatile variable, then power fails, then a second attempt to
re-execute the same computation {\em reads} the value written in the first
attempt.  Figure~\ref{fig:code_demo_incosistency} illustrates how state can
become inconsistent in an intermittent execution using cyclic redundancy check
(CRC) code from MIBench~\cite{hicks_mibench2_2016}. The code computes CRC for
an $\texttt{nBytes}$ byte message $\texttt{msg}$ with remainder
$\texttt{Rmnd}$. Even with programmer defined checkpoints (i.e., line 3 in
Fig.~\ref{fig:crc_example}) of registers, stack, and non-volatile variables
(labeled \texttt{NV}), the code may compute  $\texttt{data}$ incorrectly.  The
code includes a write after read (WAR) hazard of variable $\texttt{Rmnd}$.  WAR
hazard is caused by consecutive write and read in-between two consecutive
checkpoints using $\texttt{checkpoint()}$ function: $\texttt{data}$ variable is
written \emph{before} (line 10 in Fig.~\ref{fig:crc_example_war})
$\texttt{Rmnd}$ is read (line 6 in Fig.~\ref{fig:crc_example_war}), or more
specifically---updated before a new execution loop.

As the figure illustrates, checkpoint-based systems risk violating the
consistency of volatile and non-volatile memory~\cite{dino}. To maintain memory
consistency, versioning systems~\cite{dino,ratchet} {\em version} a subset of
non-volatile variables with the checkpointed execution context. Task-based
systems~\cite{chain,alpaca}, which we focus on this work and discuss in detail
next, ensure state is consistent by using programmer, compiler, and runtime
support.

\subsection{Task-based Intermittent Computing}
\TODO{This sequential / modular split seems pretty odd to me.  I think we can just talk about tasks in the sense of ratchet, alpaca, and chain.  Anything with static (programmer-written or compiler inserted) boundaries applies}
Intermittent execution models combatting WAR can be classified into: \emph{sequential execution model} (which historically formed a baseline for this stream of research) and \emph{modular execution model} (a potent alternative, on which we base the design decisions of \sys).
)

>>>>

In this section, we provide background on energy harvesting systems and the intermittent software execution model. We then discuss the key limitations of existing task-based execution models for intermittent computing that \sys
addresses.

@@

Energy harvesting devices operate using energy extracted from their environment, from sources such as radio waves (RF) and solar energy. Energy-harvesting devices need not use a tethered power supply or battery and instead, typically collect energy into a (super-)capacitor, operate when sufficient energy has accumulated, and upon depleting the stored energy, turn off and wait to collect more energy.

Batteryless operation has a number of important advantages, making intermittent computing an important research domain. Supplying power to billions~\cite{gartner_iot} of embedded computers using batteries is not sustainable. The European Commission estimates that more than 160 kilotons of consumer batteries enter the European Union annually~\cite{eu_batteries_2016}. Batteries are an environmental risk, are fragile, are limited in their number of charge/discharge cycles, and may require costly physical maintenance that is difficult or impossible deployed (e.g. in space~\cite{kicksat}). By contrast, super-capacitors are durable, promising millions of charge/discharge cycles~\cite[Sec. I]{ongaro_pwre_2012}. The main limitation in moving to capacitor-based energy storage is that capacitor energy density---and consequently operating discharge time---is orders of magnitude less than a battery.

%Given current technology development, battery-less systems are best suited for very long-term sensing and monitoring where access to recharge is either prohibitive or impossible. These include battery-less image capture and processing~\cite{naderiparizi_rfid_2015}, animal monitoring~\cite{thomas_jbcs_2012} or implantable~\cite{rodriguez_tbcs_2015} and digestible~\cite{nadeau_naturebio_2017} sensors.

@@

\subsection{Intermittent Computing}
\label{sec:background_consistency}

Energy-harvesting devices execute software according to the {\em intermittent execution model}~\cite{dino,chain,alpaca,ratchet}. Physically, a device charges until a threshold, operates briefly until its energy is depleted, shuts down recharges, and repeats the cycle. Software running on the device executes {\em intermittently} because energy is only available between when the capacitor reaches its threshold and when it is depleted. An intermittent execution is composed of operating periods interspersed with power failures. The frequency of power failures in an intermittent execution depends on the size of the device's energy storage buffer: a larger buffer allows longer operating periods. Energy-harvesters provide input power that is orders of magnitude less than a device's operating power, making recharging negligible during operation.

@@

Data may become inconsistent if an attempt to execute some computation {\em writes} to a non-volatile variable, then power fails, then a second attempt to re-execute the same computation {\em reads} the value written in the first
attempt. Figure~\ref{fig:code_demo_incosistency} illustrates how state can become inconsistent in an intermittent execution using cyclic redundancy check (CRC) code from MIBench2~\cite{hicks_mibench2_2016}. The code computes CRC for an $\texttt{nBytes}$ byte message $\texttt{msg}$ with remainder $\texttt{Rmnd}$. Even with programmer defined checkpoints (i.e., line 3 in Fig.~\ref{fig:crc_example}) of registers, stack, and non-volatile variables (labelled \texttt{NV}), the code may compute $\texttt{data}$ incorrectly. The code includes a write after read (WAR) hazard of variable $\texttt{Rmnd}$. WAR hazard is caused by consecutive write and read in-between two consecutive checkpoints using $\texttt{checkpoint()}$ function: $\texttt{data}$ variable is written \emph{before} (line 10 in Fig.~\ref{fig:crc_example_war}) $\texttt{Rmnd}$ is read (line 6 in Fig.~\ref{fig:crc_example_war}), or more specifically---updated before a new execution loop.

As the Figure~\ref{fig:code_demo_incosistency} illustrates, checkpoint-based systems risk violating the consistency of volatile and non-volatile memory~\cite{dino}. To maintain memory consistency, versioning systems~\cite{dino,ratchet} {\em version} a subset of non-volatile variables with the checkpointed execution context. Task-based systems~\cite{chain,alpaca}, which we focus on this work and discuss in detail next, ensure state is consistent by using programmer, compiler, and runtime support.

\subsection{Task-based Intermittent Computing}
\label{section:background_task_computing}

%\TODO{This sequential/modular split seems pretty odd to me. I think we can just talk about tasks in the sense of ratchet, alpaca, and chain.  Anything with static (programmer-written or compiler inserted) boundaries applies}

Intermittent execution models combatting WAR can be classified into: \emph{sequential execution model} (which historically formed a baseline for this stream of research) and \emph{modular execution model} (a potent alternative, on which we base the design decisions of \sys). \todo{Remove division per sequential/modular split}{Przemek}

### 2017-07-31 paper/evaluation.tex
placeholder for figures in evaluation

Expected results:
\begin{itemize}
	\item For all apps: show amount of task merged, time to completion, overhead (in terms of what?), energy consumed. All experiments for fixed power, at 3 distances with wireless power, compared against Chain 
\end{itemize}

@@

Expected results:
\begin{itemize}
	\item page swapping cost in terms of execution time: compare against Chain and full double buffering approach; show number of page swaps---do the experiment for all applications
\end{itemize}

>>>>

Expected results:
\begin{itemize}
	\item performance for each application comparing bars for Chain, no coalescing/hand-tasked, coalescing/hand-tasked, no-coalescing/compiler, coalescing/compiler. Performance depicted, normalized to one of the configurations (probably Chain). This plot will cover most of the eval dimensions that we care about.
	\item Sensitivity plot---show amount of task merged, time to completion, overhead (in terms of task-size/cap-size sweep (remember we can vary either with approximately the same effect)), energy consumed. All experiments for fixed power, at 3 distances with wireless power, compared against Chain for 3 task/cap size configurations
\end{itemize}

%Point for decision after main figure is ready

%Pull the compiler stuff into its own plot that compares compiler to hand-tasked, maybe only for the coalescing fig.  This reduces bars in plot 1, but runs the risk of making the compiler sound secondary, which we do not want to do.

%*Report only Chain, coalescing/hand-tasked, coalescing/compiler in the main plot.  This says that coalescing is the "default configuration" for VIPER. Later, in a second figure, we compare coalescing to no coalescing for hand-tasked and compiler as a characterization plot.

@@

Expected results:
\begin{itemize}
	\item Sensitivity plot---normalized page swapping cost ("bathtub" curve) in terms of execution time: compare against Chain and full double buffering approach; show number of page swaps---do the experiment for all applications
\end{itemize}

### 2017-07-31 paper/intro.tex

There are two main approaches for dealing with data inconsistency for
intermittently-powered devices: (i) programming and execution
models~\cite{dino,ratchet,chain,alpaca} and (ii) architecture
mechanisms~\cite{hicks_isca_2017,idetic,nvp,tictpl}. While both approaches represent
progress, each has critical limitations. New architectures require costly
hardware changes and are inapplicable to today's embedded off-the-shelf
systems~\cite[Fig.  3]{hicks_isca_2017}~\cite[Fig. 9]{nvp}. New programming
models and compilers force programmers to change behavior~\cite{chain,ratchet},
inhibiting adoption.  New memory models may require more accesses to
non-volatile memory (for multi-versioning)~\cite{dino,chain} or may preclude
the use of volatile memory~\cite{ratchet}.  Effective use of both volatile and
non-volatile memory is important for efficiency, because volatile memory has
higher access energy and latency than SRAM~\cite[Fig. 4]{nvp}, and generality,
because devices often have much more FRAM than SRAM (e.g., 64 times
more~\cite{wolverine}). Hardware is effective, but inapplicable to existing
devices, making programming and execution models essential to the success of
intermittent computing. 
%Despite the limitations, we conjecture that programming
%and execution models will be a {\em de facto choice for sustaining data
%consistency} of intermittently-powered devices. Simply put, programming model
%does not require any hardware investment and is dependent only upon the
%programmer's skills.

Recent work~\cite{alpaca,chain} proposed {\em task-based} programming and
execution models in which the programmer statically decomposes their program
into a collection of tasks.  Tasks can include arbitrary computation and, upon
completion, are guaranteed to have executed {\em atomically}, despite
arbitrarily-timed power failures. 
%
The programmer explicitly expresses task-to-task control flow, which may be
conditionally input-dependent~\cite[Fig. 4]{chain}. An important programming
challenge presented by a task-based model is that the length of a software task's execution
is limited by the fixed, total amount of energy that a device can buffer in
hardware.  A task's code is static, but the duration of its execution
may be input-dependent and is difficult to predict.
%
Assuming that input power is negligible compared to operating
power~\cite{wisp}, a task will never be able to complete if its execution
consumes more energy than the system can buffer, creating the potential for
long tasks to be irreparably non-terminating.  On the other hand, transitioning
from one task to another imposes a run time cost and excessively small tasks
are inefficient.

@@

\TODO{This is the main pitch statement.  we should all agree that this is what
\sys does.} In this work we develop {\bf \sys}: a new programming and execution
model that efficiently executes tasks and avoids non-termination across a range
of energy buffer sizes, without burdening the programmer.
%
To accomplish this goal, \sys introduces three new capabilities: {\em automatic
task compilation} that compiles arbitrary C code into tasks, {\em dynamic task
coalescing} that efficiently executes tasks without exceeding available energy,
and {\em volatile memory virtualization} that efficiently leverages both
volatile and non-volatile memory.  
%
\sys's compiler transforms arbitrary C code into a graph of tasks that reflects
the original program's control- and data-flow constraints.  The compiler
identifies data shared by multiple of the newly created tasks and instruments
reads and writes of those data so that \sys's memory virtualization mechanism
keeps those data consistent.
%
\sys's memory virtualization mechanism uses SRAM as working memory, that \sys
dynamically populates with pages of data from FRAM on demand during a task's
execution.  When a task ends, \sys commits dirty pages to FRAM using efficient
DMA block copies, ensuring task atomicity and memory consistency.
%
\sys's task coalescing mechanism executes multiple statically defined tasks
together as a single dynamic task. A programmer can specify small tasks that
will execute on a device with a small energy buffer. Coalescing tasks allows
running the same tasks on a device with a larger energy buffer, while avoiding
run time overheads associated with ending one task and beginning another.  
%

%~\cite{kistler_micro_2006} This reference will confuse people familiar with DMA.  DMA is well known I think we should cite the MCU datasheet, or use no cite. 
%
\todo{articulate this last point better}{Przemek}

To summarize, this work makes the following contributions:
%
\TODO{I think these were all much too long, given the depth of the intro and background already.  I am cutting to the barest minimum.}
\begin{enumerate}
%
\item {\bf Task coalescing mechanism:} \sys's core contribution is the introduction of dynamic task coalescing mechanism. It allows the programmer or a compiler to intuitively decompose the program into small tasks that amortize fixed per-task overheads, yet present no risk of exceeding device energy capacity. As such a decomposition executes, \sys~{\em coalesces dynamically} consecutive tasks. Coalescing elides the commits of coalesced tasks by buffering multiple tasks' updates in an FRAM commit buffer. Periodically, as the span of the coalesced task grows, \sys ends coalescing and commits the state of the coalesced task. If a power failure interrupts a sequence of coalesced tasks, \sys adaptively reduces the number of tasks in that sequence that it will coalesce, committing sooner in future executions. Consequently, \sys's coalescing mechanism allows a program to execute efficiently across a range of energy buffer sizes, avoiding transition overheads in larger buffers, and ensuring progress in smaller buffers.
%
\todo{Provide numerical results for this section}{Przemek}

\item {\bf Compiler-support automatic task generation:} second \sys's core contribution is the ability to split standard C programs into \sys's compatible tasks, effectively circumventing the need for labour-intensive manual task generation of the state of the art solutions~\cite{chain,alpaca}.
%
\todo{Elaborate on this more and provide numerical results}{Przemek}
%
\item {\bf Evaluation of data swapping strategies:} We propose and explore two data swapping strategies for intermittently-powered systems based on FRAM/SRAM architecture. The first strategy is {\em demand paging}, which swaps the SRAM page with a new page from FRAM, buffering the swapped-out page until commit. The second strategy is {\em buffered direct access}, which directly reads and writes FRAM relying on dynamic double-buffering to ensure memory is consistent at
commit. 
%
\todo{Once paging is fully implemented--update the paragraph}{Przemek}
\todo{Provide numerical results for this section}{Przemek}
%
\item {\bf Extensive \sys evaluation:} We evaluated \sys end-to-end on a collection of existing benchmark programs~\cite[Sec. 5]{chain} and new ones (including realistic case study of battery-less embedded sound detector) in a laboratory environment, and compare directly to state-of-the art task-based runtime~\cite{chain}, showing that \sys has high performance and can flexible target a variety of platforms without recompiling the code. 
%
\todo{add detailed text about evaluation results}{Przemek}
\todo{decide if we want to make this a contribution paragraph or not (then, put it as a closing paragraph)}{Brandon}
%
\end{enumerate}

>>>>

There are two main approaches for dealing with data inconsistency for intermittently-powered devices: (i) programming and execution models~\cite{dino,ratchet,chain,alpaca} and (ii) architecture
mechanisms~\cite{hicks_isca_2017,idetic,nvp,tictpl}. While both approaches represent progress, each has critical limitations. New architectures require costly hardware changes and are inapplicable to today's embedded off-the-shelf systems~\cite[Fig. 3]{hicks_isca_2017}~\cite[Fig. 9]{nvp}. New programming models and compilers force programmers to change behavior~\cite{chain,ratchet}, inhibiting adoption.  New memory models may require more accesses to non-volatile memory (for multi-versioning)~\cite{dino,chain} or may preclude
the use of volatile memory~\cite{ratchet}. Effective use of both volatile and non-volatile memory is important for efficiency, because volatile memory has higher access energy and latency than SRAM~\cite[Fig. 4]{nvp}, and generality, because devices often have much more FRAM than SRAM (e.g., 64 times more in~\cite{wolverine}). Hardware is effective, but inapplicable to existing devices, making programming and execution models essential to the success of intermittent computing.

Recent work~\cite{alpaca,chain} proposed {\em task-based} programming and execution models in which the programmer statically decomposes their program into a collection of tasks. Tasks can include arbitrary computation and, upon completion, are guaranteed to have executed {\em atomically}, despite arbitrarily-timed power failures. The programmer explicitly expresses task-to-task control flow, which may be conditionally input-dependent~\cite[Fig. 4]{chain}. An important programming challenge presented by a task-based model is that the length of a software task's execution
is limited by the fixed, total amount of energy that a device can buffer in hardware.  A task's code is static, but the duration of its execution may be input-dependent and is difficult to predict. Assuming that input power is negligible compared to operating power~\cite{wisp}, a task will never be able to complete if its execution consumes more energy than the system can buffer, creating the potential for long tasks to be irreparably non-terminating. On the other hand, transitioning from one task to another imposes a run time cost and excessively small tasks are inefficient.

@@

\todo{This is the main pitch statement: we should all agree that this is what \sys does}{All} In this work we develop {\bf \sys}: a new programming and execution model that efficiently executes tasks and avoids non-termination across a range of energy buffer sizes, without burdening the programmer. To accomplish this goal, \sys introduces three new capabilities:

\begin{itemize}
	\item {\bf Automatic task compilation:} \sys's compiler transforms arbitrary C code into a graph of tasks that reflects the original program's control- and data-flow constraints. The compiler identifies data shared by multiple of the newly created tasks and instruments reads and writes of those data so that \sys's memory virtualization mechanism keeps those data consistent. \sys complier effectively circumvents the need for labour-intensive manual task generation of the state of the art solutions~\cite{chain,alpaca}. \todo{Elaborate on this more and provide numerical results}{Przemek}
	%
	\item {\bf Dynamic task coalescing:} \sys's memory virtualization mechanism uses SRAM as working memory, that \sys dynamically populates with pages of data from FRAM on demand during a task's execution. When a task ends, \sys commits dirty pages to FRAM using efficient DMA block copies, ensuring task atomicity and memory consistency. \todo{Once paging is fully implemented--update the paragraph}{Przemek}\todo{Provide numerical results for this section}{Przemek}
	%
	\item {\bf Volatile memory virtualization:} \sys's task coalescing mechanism executes multiple statically defined tasks together as a single dynamic task. A programmer can specify small tasks that will execute on a device with a small energy buffer. Coalescing tasks allows running the same tasks on a device with a larger energy buffer, while avoiding run time overheads associated with ending one task and beginning another.\todo{Provide numerical results for this section}{Przemek}
\end{itemize}

We evaluated \sys end-to-end on a collection of existing benchmark programs~\cite[Sec. 5]{chain} and new ones (including realistic case study of battery-less embedded sound detector) in a laboratory environment, and compare directly to state-of-the art task-based runtime~\cite{chain}, showing that \sys has high performance and can flexible target a variety of platforms without recompiling the code. \todo{add detailed text about evaluation results}{Przemek}

%\todo{I think these are all much too long, given the depth of the intro and background already. I suggest we cut to the barest minimum.}

%\sys's core contribution is the introduction of dynamic task coalescing mechanism. It allows the programmer or a compiler to intuitively decompose the program into small tasks that amortize fixed per-task overheads, yet present no risk of exceeding device energy capacity. As such a decomposition executes, \sys~{\em coalesces dynamically} consecutive tasks. Coalescing elides the commits of coalesced tasks by buffering multiple tasks' updates in an FRAM commit buffer. Periodically, as the span of the coalesced task grows, \sys ends coalescing and commits the state of the coalesced task. If a power failure interrupts a sequence of coalesced tasks, \sys adaptively reduces the number of tasks in that sequence that it will coalesce, committing sooner in future executions. Consequently, \sys's coalescing mechanism allows a program to execute efficiently across a range of energy buffer sizes, avoiding transition overheads in larger buffers, and ensuring progress in smaller buffers.

%We propose and explore two data swapping strategies for intermittently-powered systems based on FRAM/SRAM architecture. The first strategy is {\em demand paging}, which swaps the SRAM page with a new page from FRAM, buffering the swapped-out page until commit. The second strategy is {\em buffered direct access}, which directly reads and writes FRAM relying on dynamic double-buffering to ensure memory is consistent at commit.
